---
title: "Basics of Coding in R"
author: "Tiep Nguyen"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

R is the primary language used in coding for RStudio. There are two different types of file that support operations in R. 

## Basic Operations

R supports basic math operations with the common ones listed below.

```{r}
# Addition
2 + 3

# Subtraction
5 - 2

# Multiplication
4 * 3

# Division
10 / 2

# Exponentiation
2^3
```

# Variables

Variables are used to store specific values to keep track of. Variables can be initialized with either the <- or -> operator, such as
```{r, eval = FALSE}
  count <- 49
  "Heldi" -> current_name
  
```
Variables are useful for calculations, especially for values that change such as for example

```{r, eval = FALSE}

price <- 30
items_sold <- 60
revenue <- price * items_sold
```

Not only that, variables can be initialized to multiple times.

```{r, eval = FALSE}
total_sheep <- 50
new_sheep <- 10
total_sheep <- total_sheep + new_sheep # this reassigns total sheep to be its new value plus the old value
```
# Naming Conventions for Variables
Variables should consist of only alphanumeric characters and '_' characters. The style of variable names depends on the lab and researcher, but it's important to be consistent. Here are some common naming conventions

my_variable_name - snake_case

My_Variable_Name - Title_Case

MyVariableName - PascalCase

myVariableName - camelCase


Generally I prefer snake_case for writing in R, but as long as you are consistent with your naming conventions you are good to go.

# Data Types

Variables hold specific data types. Every value a variable hold has its own data type. Unlike in other programming languages, variables can hold any value. We'll learn more about data types in the following pages

# Functions

Functions are reusable blocks of code that allow you to do the same actions repeatedly. For example, the sqrt() function returns the square root of the number.

Functions have a list of named arguments and default arguments. Arguments, which are basically your input values, need to be of a specific data type to work. Lets take the round function for example

```{r}
print(round(3.145)) # this prints "3"
```

However, the round function actually has two parameters, "x", which is our numeric vector (basically a number) and "digits". By default, digits = 0, which means it rounds to the 0 digits after the decimal point. We can put in the parameter either in order or by name.

```{r}
print(round(3.145, 3)) # this prints "3.145"
print(round(3.145, digits=2)) # this prints "3.14"
print(round(x=3.145, digits=1)) # this prints "3.1"
print(round(digits=1, x=3.145)) # this also prints "3.1"
```
We can also define our own functions. We can define a function by assigning a value to a variable, like so:
```{r}
custom_round <- function(x, digits = 0) {
  # Scale X up.
  factor <- 10^digits
  scaled_x <- x * factor
  
  # generates two possible values
  floorX <- floor(scaled_x)
  ceilX <- ceiling(scaled_x)
  
  result <- -1 # initializes result
  #finds which value is closer to X, and picks that one.
  
  if (abs(scaled_x - floorX) <= abs(scaled_x - ceilX )) { # we replicate R's implementation by rounding down when it is 0.5
    result <- floorX
  }
  else {
    result <- ceilX
  }
  
  # Scale back to the original number
  result <- result / factor
  
  return(result)
}
```
```{r}
print(custom_round(3.145, 3)) # this prints "3.145"
print(custom_round(3.145, digits=2)) # this prints "3.14"
print(custom_round(x=3.145, digits=1)) # this prints "3.1"
print(custom_round(digits=1, x=3.145)) # this also prints "3.1"


```

Note that functions are stored in a variable with their parameter names as arguments. Certain functions require arguments, like how round always requires a value in x while digits is a default argument because it chooses 0 if not specified.

Functions can also have variables inside them, like resultX and ceilX. However, these variables are limited in scope to that function, meaning that the function and other commands outside of that function can affect them. While functions can affect variables outside of the function or its parameters, it is not recommended to do so.

# Library Imports

Unfortunately, writing functions takes time and energy. Even a simple function like round required about 10 lines of code for a naive implementation. Fortunately, R allows us to write premade functions and export them so that other people can use them.

Libraries are essentially bundles of functions that can be imported into projects. Generally libraries fulfil specific niches that not every project needs. Some examples of common R libraries include
  Dplyr, which is used for data manipulation
  ggplot, which is used for data visualization (e.g graphs)
  tidyverse, a collection of packages designed to provide easy out of the box functionality for your use.

Libraries can be installed by using the command
```{r, eval = FALSE}
install.packages("your_package_name")
library("your_package_name")
```
This adds the package to the environment for you to run in the command line or in scripts.

Note that when you install packages in a script, every time you run that script it demands an environment restart in order to avoid duplicating that existing package. Theres two ways of avoiding this
  Run your code without the imports once you've imported the packages
  Use requireNameSpace to see if it is already installed, like so
```{r, eval = FALSE}
    if (!requireNamespace("your_package_name", quietly = TRUE)) { # checks if the package is in the namespace.
      install.packages("your_package_name")
    }
  library("your_package_name")

```